Predicting FOMC Actions using NLP; Assesing the impact of matrix sparsity and regularization

### 2019-05-13 presentation/paperv2.tex
paperv2

% AMS packages:
\usepackage{amsmath, amsthm, amsfonts}

@@

%-----------------------------------------------------------------
\title{Predicting FOMC Actions using ML and NLP}
\author{
        Jeremy Lao - jjl359 \\
        NYU Computer Science \\
            \and
        John Reynolds - jr4716 \\
        NYU Computer Science \\
}

\begin{document}
\maketitle

\abstract{\textit{The Federal Open Market Committee (FOMC) meets throughout the year to set the target Federal Funds rate.  The Federal Funds rate is one of the primary monetary policy tools, and it impacts the cost of borrowing money globally.  This paper explores ML and NLP methods to predict the potential outcome of an upcoming FOMC using past FOMC statements and meeting minutes to train ML and NLP models.}}

\section{Introduction}
In this paper we methods in Natural Language Processing (NLP) and Machine Learning (ML) to predict Federal Open Market Committee (FOMC) rate actions (hold or change) using text from FOMC meeting minutes, Board of Governors speeches, and FOMC post-meeting statements.  Since this is work in sentiment analysis we created a simulator to gain insight into the sensitiviy of detecting sentiment by randomly generating words with from a mix of distributions containing hawkish or dovish(positive) words, negative words, and commonly used words.  That simulator showed accuracy could hit levels of 100 percent given enough training data and high percentage of sentiment (i.e., positive)  words to overall text.

@@

\section{Objective}

The aim of our research is to determine whether ML and NLP methods can be used to predict FOMC action (no change or a change in the base rate).  Our models approached this question in two steps:

\begin{enumerate}

 \item We test the accuracy of ML models trained on a random mixture of FOMC minutes, speeches, and statements on a test set of FOMC documents.  In this stage, we are determining the viability of ML methods in predicting FOMC action on test documents, and we will evaluate the model accuracy with precision and recall.  The models are: 
 \begin{itemize}
    \item Naive Bayes
    \item Logistic Regression 
    \item SVM
    \item Logistic Lasso
 \end{itemize}
 \item We create a predictive model that is trained on documents that are temporally released before the test document.  Typically we will test with FOMC minutes releases or Board of Governor speeches that are temporally released after the training documents. 
 \item  To determine the appropriate feature vector size, for example, concatenate the documents associated with a meeting or treat each document separately, we simulated how sentiment analysis performs as the feature vector becomes larger. 
 \item  We applied Latent Dirichlet Allocation to the entire corpus of speeches and assumed $k=3$ latent topics to explore whether LDA captures the Federal Reserve's mandates of: maximum employment, stable prices, and financial stability. 
\end{enumerate}

@@

We primarily utilized three sources for our data, they are outlined below. \vertSpace

The time frame of our study covered 2000 to 2018.  The primary source of our textual data were FOMC statements, FOMC meeting minutes, and Board of Governors speeches from \url{www.federalreserve.gov}.  For FOMC statements between 2000 and 2008 we utilized Stanford's FOMC corpus.  The Stanford FOMC corpus only has text prior to 2008 and it does not have Board of Governor speeches.  \vertSpace

The data for the Federal Funds target rate and target range (post 2009) is from FRED St. Louis.  FRED offers a wealth of economic data and information to promote economic education and enhance economic research. FRED is updated regularly and allows 24/7 access to regional and national financial and economic data.  \vspace{5mm}

\noindent Below is a table of the FOMC document metadata:

@@

\subsection{Scraping and Text Pre-Processing}

\subsection{Models}

\subsubsection{Latent Dirichlet Allocation}

@@

\subsubsection{Supervised Models}

We utilized the following models in our research:

  \begin{itemize}

    \item Naive Bayes

      $$P(A|x_1,x_2,...,x_n) = p(A)\prod_{i=1}^n p(x_i|A)$$

    \item Logistic Regression

     $$A = \prod_{i=1}^n p(x_i)^{y_i}(1-p(x_i))^{1-y_i}$$

    \item Logistic Lasso
     $$A = \prod_{i=1}^n p(x_i)^{y_i}(1-p(x_i))^{1-y_i}+ L1$$
    \item SVM

     $$\left [\frac{1}{n} \sum_{i=1}^n max(0,1 - y_i(\vec{w} \dot \vec{x_i} - b)) \right] + \lambda ||\vec{w}||^2$$
    \item Neural Network
      We test a mixture of wide and deep neural nets. 
  \end{itemize}

\subsubsection{Model Choice}

To identify the optimal prediction model we developed a simulator that would randomly create a training set and select a test set of meetings that temporally came after the training set: $$E[D_t | \mbox{Statements}_{t-i}, \mbox{Minutes}_{t-j} , \mbox{Speeches}_{t-k}, \mbox{N-gram}]$$

We tested for model accuracy through averaging various iterations of each model's results, we evaluated model performance with F1 scores (precision and recall), and we empirically tested for the optimal number of Ngrams for each model.  \vertSpace

Given our binary classification objective, we expected the Logistic Regression with L1 penalty (i.e., Logistic Lasso) to perform the best.  Based on empirical knowledge, FOMC rate moves are analyzed in terms of probabilities and the upcoming meeting is judged primarily on the outcome of Fed action or no action.  \vertSpace

Despite some good properties of SVM such as, SVM generally performs better than logistic regression due to the face that SVMs are less sensitive to outliers and SVM try to maximize the margin between the closest support vectors. However, becuase SVM provides prediction of 1 or 0 and Logistic Regression provides a probability, where a probability of 0.501 is enough for the prediction of FOMC rate action, Logistic Lasso most closely resembles how an interest rate expert would view FOMC action, a probabilistic approach.  \vertSpace

\subsubsection{Feature Choice}

In our search for the optimal model, we tested different strategies for feature selection, namely different size Ngrams and document stacking (i.e., combining the minutes, statements, and speeches for a meeting into one document):

$$E[D_t | \mbox{AGG}(\mbox{Statements}_{t-i}, \mbox{Minutes}_{t-j} , \mbox{Speeches}_{t-k}), \mbox{N-gram}]$$

Document stacking is a basic way to reduce the dimensionality of each meeting's documents, and we thought it would be an appropriate basic strategy to enhance model performance based on research from [footnote]. \vertSpace

For Ngrams, based on previous research analyzing FOMC transcripts \cite{stanford} with ML models, tri-grams wiht a Naive Bayes model performed the best for classifying the outcome of an FOMC meeting.  In line with that research, we found that N-grams of greater N generally resulted in greater accuracy and lower standard deviation.  \vertSpace

\subsection{Model Performance and Results}

Based on our simulation we observed that the Logistic regression with L1 penalty (i.e. Logistic Lasso) performed the best for Ngrams between two and ten and we attribute this to the model being well suited for the binary classification problem.  We exptected Logistic Regression with L1 penalty (i.e., Logistic Lasso) to perform better than Logistic Regression without any regularization.  \vertSpace

Given that we are modelling a classification problem, we would expect the performance of the Logistic Lasso model based on the mathematical properties of Logistic Lasso more robust for predicting FOMC rate action.  Logistic regression models the probability distribution of the class label $y$ given a feature vector $x$:

$$p(y=1|x,\theta) = \sigma(\theta^T x) = \frac{1}{1+\exp(-\theta^Tx)}$$

The L1 regularized logistic regression (Logistic Lasso) is defined as:

$$\underset{\rm \theta}{\rm min} \sum_{i=1}^M -\log p(y^{(i)}|x^{(i)}; \theta) + \beta ||\theta||$$

L1 regularization has the effect of forcing feature selection, and as we observe the Logistic Lasso model performs best with Tri-grams to 7-grams while the F1 measure significantly drops with larger Ngrams.  \vertSpace

Below is a picutre of model performance which shows that the Logistic Lasso has the highest F1 $(>90\%)$

\begin{figure}[H]
\includegraphics[width=1\textwidth]{"side-by-side".pdf}
\end{figure}

\subsection{Conclusion}

@@

\bibitem{stanford} Puri, Indira \emph{Using Machine Learning to Predict Interest Rate Changes from Federal Reserve Proceedings}
\end{thebibliography}

>>>>

% AMS packages:
\usepackage[fleqn]{amsmath}
\usepackage{amsmath, amsthm, amsfonts}

@@

%----------------------------------------------------------------
\title{Predicting FOMC Actions using NLP; Assesing the impact of matrix sparsity and regularization}
\author{
        Jeremy Lao - jjl359 \\
        NYU Computer Science \\
            \and
        John Reynolds - jr4716 \\
        NYU Computer Science \\
}

\begin{document}
{\setlength{\mathindent}{0cm}
\maketitle

\abstract{\textit{The Federal Open Market Committee (FOMC) meets throughout the year to set the target Federal Funds rate.  The Federal Funds rate is one of the primary monetary policy tools, and it impacts the cost of borrowing money globally.  This paper explores ML and NLP methods, examines the impact of matrix sparsity from NLP methods and regularization from ML models, to predict the potential outcome of an upcoming FOMC using past FOMC statements and meeting minutes to train ML and NLP models.}}

\section{Introduction}
In this paper we use methods in Natural Language Processing (NLP) and Machine Learning (ML) to predict Federal Open Market Committee (FOMC) rate actions (hold or change) using text from FOMC meeting minutes, Board of Governors speeches, and FOMC post-meeting statements.  
For this analysis we used 713 separate documents (rougly 1.6 Million Words) and 148 Fed Decisions for our data set.  
Employing NLP techniques can result in large numbers of features being generated and an extremly sparse resultant feature matrix.  
In the analysis we conducted some of the feature matricies generated contained over 2 Billion elements with fractions as small as 0.0025 of the elements containing a non-zero value.  
Our analysis and techniques used were driven by the issue of feature sizes orders of magnitued larger than the label size and the extreme degrees of sparsness in the feature matrix.  
To make sure the analysis was at least plausible we created a simulator to detect positive or negative sentiment from randomly generated text and found as the number of sentiment words increased the prediction accuracy approached close to 1.0.  
For the study the analysis hones in on the mix of distributions containing hawkish or dovish(positive) words.

\subsection{git repository for project}
\url{https://www.github.com/jrrpanix/ML9}

\section{Objective}
\begin{enumerate}
\item Since its pretty easy to just blindly apply tools to a set of data we wanted to understand the 
characteristics of the output from the text processing functions (in our case CountVectorizer) in terms of number of features, matrix sparsity, matrix norm
that are generated from calls to this function.
\item Assess the effeiveness of Bayesian Approaches and Regularization on the analysis.
\item Determine whether ML and NLP methods can be used to predict FOMC action (no change or a change in the base rate).
\item Because 70$\%$ of the actions are ``no action'', make sure the prediction is not always just ``no action'' (which would give a 70$\%$ accuracy).
\end{enumerate}

The time frame of our study covered 2000 to 2018. 
We primarily utilized three sources for our data, FOMC statements, FOMC meeting minutes and Boarde of Governors speeches.
Details on datasources, extraction and scrubbing are left to the appendix to keep the focus on the methods and results.
Preparing the data was a large amount of work and we are not aware of any data sets that combine the three types of documents
and then map the documents to a Fed Action.

\subsection{Feature Modeling and Data Label Mapping}
\noindent Documents are mapped to the first rate decision on or before the document date at time $ti$
where $j=\text{Statement}$ or $j=\text{Speech}$ or $j=\text{Minutes}$
\begin{equation*}
\mbox{Document}_{j,ti} \rightarrow \mbox{Decision}_{ti+k}
 = \begin{cases}
  1, & \text{if  Decision = change}, \\
  0, & \text{if  Decision = no change}
\end{cases}
\end{equation*}

\noindent 
\textbf{Model 1: Unstacked}\\
\noindent
The \textbf{Unstacked} model maps each document to a feature vector generated by CountVector given an n-gram($n_1$, $n_2$)
where $n_1 <= n_2$ and $n_i$ are integers $>0$. Given there are approximately 713 documents, this will generate
a feature matrix with 713 rows and $K$ columns where each column will be and instance of the word sequence
generated by the n-gram.

\begin{equation*}
E[D_ti] = f(ngram(n_1, n_2)[\mbox{Document}_{j,ti-k}])
\end{equation*}

\noindent
\textbf{Model 2: Stacked}\\
The \textbf{Stacked}  model maps the set of documents which occurr after the previous decision date but on or before the
next decision date to one rate decision.  This has the effect of contatenating the documents between decisions into
one large document, but cuts down the number from 713 to 148.
where $n_1 <= n_2$ and $n_i$ are integers $>0$. Given there are approximately 713 documents, this will generate
a feature matrix with 713 rows and $K$ columns where each column will be and instance of the word sequence
generated by the n-gram.

\begin{equation*}
E[D_{ti}] = f(ngram(n_1, n_2)[\mbox{Document}_{j1,ti-k1},{Document}_{j2,ti-k2},...]) \text{, } ti-1 < ti-ki < ti+1
\end{equation*}

\subsection{Matrix Sizes and Sparsity from CountVector}
\noindent For this study we used \textbf{CountVector} (\textbf{TfiDf} can also be used, but it did not alter the conclustions of this study).
When are n-gram is applied to corpus it combinatorics diticates a large number of features will be generated and for a given feature, most of the column entries will
be zero, resulting in extremly sparse matricies.  The table below gives a measure of $\mbox{sparsity} = \frac{non-zero elements}{total elements}$ and 
the resulting number of elements in each n-gram generated matrix in the billions. In general stacking reduces the sparsity by a factor of 4 and
the matrix size by a factor of 5.\\

\noindent \begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline
 \multicolumn{5}{|c|}{CountVector Matrix Size} \\
 \hline
 \multicolumn{1}{|c|}{} &
 \multicolumn{2}{|c|}{Unstacked} &
 \multicolumn{2}{|c|}{Stacked}\\
 \hline
 N-gram & sparsity & size(billions) & sparsity & size(billions)\\
 \hline
       1:1& 0.03628801& 0.008561&  0.10635940&   0.001757 \\
       1:2& 0.00544312& 0.170058&  0.02081416&    0.036107\\
       1:3& 0.00368457& 0.439898&  0.01508941&     0.092869\\
       3:5& 0.00238128& 0.900721&  0.01059293&     0.173040\\
       4:7& 0.00224699& 1.332560&  0.00996789&     0.271117\\
      5:10& 0.00218455& 1.995502&  0.00972276&     0.402041\\
      8:10& 0.00213633& 1.019811&  0.00955894&     0.198627\\
     10:15& 0.00210643& 2.097187&  0.00945586&     0.398312\\
     20:20& 0.00205055& 0.347043&  0.00931157&     0.074799\\
 \hline
\end{tabular}

\section{Conclusion}

\appendix
\section{Appendix}

@@

@@

\subsubsection{Data Sources}
The time frame of our study covered 2000 to 2018.  
The primary source of our textual data were FOMC statements, FOMC meeting minutes, and Board of Governors speeches from \url{https://www.federalreserve.gov}.  
For FOMC statements between 2000 and 2008 we utilized Stanford's FOMC corpus.  
The Stanford FOMC corpus only has text prior to 2008 and it does not have Board of Governor speeches.
The data for the Federal Funds target rate and target range (post 2009) is from FRED St. Louis.  
FRED offers a wealth of economic data and information to promote economic education and enhance economic research. 
FRED is updated regularly and allows 24/7 access to regional and national financial and economic data.

\subsubsection{Scraping and Text Pre-Processing}

\subsubsection{Meta-Data}
\noindent Below is a table of the FOMC document metadata:

@@

\subsection{Latent Dirichlet Allocation}

@@

@@

\bibitem{stanford} Puri, Indira \emph{Using Machine Learning to Predict Interest Rate Changes from Federal Reserve Proceedings}

\bibitem{nyu} Ranganath, Rajesh  \emph{NYU - CSCI-GA.2565 ML Lecture Slides Spring 2019}
\end{thebibliography}
